---
title: "Exploratory data analysis using the tidyverse and ggplot2"
format: 
 revealjs:
   smaller: true
   theme: default
   slide-number: true
   chalkboard: 
      buttons: false
   preview-links: auto
   logo: ../images/LSHTM_Logo_Black.png
   include-in-header:
     - text: |
         <style>
         .reveal .slide-logo {
           max-height: unset;
           height: 100px;
         }
         </style>
---

```{r, include=FALSE}
library(mosaicData)
library(tidyverse)
library(gapminder)
```

# Introduction

## What is the tidyverse?

A suite of R packages to make working with data as easy as possible (Wickham 2020), including: 

-  ggplot2: for plotting data 
-  dplyr: for manipulating data frames 
-  tidyr: for making data tidy 
-  forcats: for manipulating factor variables 
-  magrittr: for easy chaining of commands 

(Wickham and Grolemund 2017; Wickham et al. 2019)

# Summarising data

## Summarising data {.smaller}

-   Summary statistics one of most common data analysis tasks
-   Consider the `Gestation` data from `mosaicData` 

Birth weight, date, and gestational period collected as part of the Child Health and Development Studies in 1961 and 1962. Information about the baby's parents --- age, education, height, weight, and whether the mother smoked is also recorded (Nolan and Speed 2001).

-   We will use some functions from dplyr to choose, group and summarise data
-   `verb(.data, ...)` applies a dplyr verb to a data frame

## Summarising data

-   count how many babies in data set

```{r, echo=TRUE, eval=TRUE}
library(mosaicData) 
library(tidyverse) 

data(Gestation) 
count(Gestation) # just like nrow() 
```

## Summarising grouped data

-   Can also count for a given grouping variable

```{r, echo=TRUE, eval=TRUE}
count(Gestation, race) # nrow can't do this 
```

## Summarising data

-   The summarise function allows us to calculate summary statistics of a variable
-   Can (and should) give names to summary columns
-   Calculate the mean birth weight in the data set

```{r, echo=TRUE, eval=TRUE}
summarise(Gestation, wt_mean = mean(wt)) 
```

## Summarising data

-   We can calculate multiple summaries at once

```{r, echo=TRUE, eval=TRUE}
summarise(Gestation,           
          Mean = mean(wt), 
          SD   = sd(wt), 
          Low  = quantile(wt, 0.025), 
          High = quantile(wt, 0.975)) 
```

## Summarising grouped data {.smaller}

-   We can group the rows in our data and calculate summaries for each group
-   `group_by` lets us pass variable names to set the structure
-   Row order is maintained

```{r, echo=TRUE, eval=TRUE}
Gestation_grouped_by_race <- group_by(Gestation, race) 
Gestation_grouped_by_race 
```

## Summarising grouped data

-   `summarise()` respects the grouping structure

```{r, echo=TRUE, eval=TRUE} 
summarise(Gestation_grouped_by_race,           
          Mean = mean(wt), 
          SD   = sd(wt), 
          Low  = quantile(wt, 0.025), 
          High = quantile(wt, 0.975)) 
```

## Summarising data

-   We can summarise multiple variables (`.vars)` with the same function(s) (`.funs`)

```{r, echo=TRUE, eval=TRUE}
summarise_at(Gestation,  
             .vars = vars(gestation, wt, age), 
             .funs = mean, na.rm = T) 
```

## Summarising data

-   We can even do multiple summaries across multiple variables by specifying
    -   which `.vars` we want to summarise, and
    -   a list of which z we wish to summarise with

```{r, echo=TRUE, eval=TRUE}
summarise_at(Gestation,  
             .vars = vars(gestation, wt, age),
             .funs = list(mean = mean,sd = sd), 
             na.rm = T) 
```

## Summarising grouepd data

-   All `summarise\*()` respects the grouping structure

```{r, echo=TRUE, eval=TRUE}
summarise_at(Gestation_grouped_by_race, 
             .vars = vars(gestation, wt, age), 
             .funs = list(mean = mean), na.rm = T) 
```

# Rows and columns

## Operating on columns

-   create/modify/delete columns with dplyr's `mutate()`
-   e.g. relabelling `race` so words start with a capital

```{r, echo=TRUE, eval=TRUE}
Gestation <- mutate(Gestation, race = str_to_title(race)) 
count(Gestation, race) 
```

## Choosing columns {.smaller}

-   For one reason or another we may want to select only certain columns of our data frame

```{r, echo=TRUE, eval=TRUE}
head(Gestation, 1)
head(select(Gestation, race, wt, number), 1)
```

## Choosing and renaming columns

-   We can also rename columns on the fly as we select them

```{r, echo=TRUE, eval=TRUE}
select(Gestation, 
       Race                = race,  
       `Birthweight (oz)`  = wt,                                   
       Cigs.smoked         = number) 
```

## Choosing and renaming columns

-   Alternatively we can rename columns without worrying about failing to select columns we haven't renamed

```{r, echo=FALSE, eval=TRUE}
rm(Gestation)
data(Gestation)
Gestation <- mutate(Gestation, race = str_to_title(race)) 

```

```{r, echo=TRUE, eval=TRUE}
names(Gestation)

Gestation <- rename(Gestation, Cigs.smoked = number)

names(Gestation)
```

## Choosing rows {.smaller}

-   The dplyr equivalent of subset is filter
-   Takes a logical statement and does non-standard evaluation of variable names
-   `filter(data, A & B)` the same as `filter(data, A, B)`

```{r, echo=FALSE, eval=TRUE}
rm(Gestation)
data(Gestation)
Gestation <- mutate(Gestation, race = str_to_title(race)) 

```

```{r, echo=TRUE, eval=TRUE}
Gestation2 <- select(Gestation, 
                     Race                = race,  
                     `Birthweight (oz)`  = wt,                                   
                     Cigs.smoked         = number) 
filter(Gestation2, Race == "White", Cigs.smoked == "never") 
```

## Choosing rows {.smaller}

-   `slice\*()` functions allow you to select rows based on their properties, e.g. which babies have lowest birth weight overall and in each race group?

```{r, echo=TRUE, eval=TRUE}
slice_min(Gestation2, `Birthweight (oz)`) 
slice_min(group_by(Gestation2, Race), `Birthweight (oz)`) 
```

# Reshaping data frames

## Long and wide tidy data

-   Tidy data frames consist of a number of observations (rows) of variables (columns), they can be either wide or long (Wickham 2014)
-   We can pivot between wide and long format with tidyr's `pivot\_\*()` functions
-   Pivoting to a longer data frame helps put data in key-value pairs, useful for ggplot2
-   The value is the value of the named variable for a given id

## Long and wide tidy data

To make this pivot, we specify 

-   which cols are to be converted from being `k` columns of length `n` to one column of length `n×k`
-   the `names_to` column name, which contains the names of the pivoted columns 
-   the `values_to` name of the column containing the value of each variable for each `id`

## Long and wide tidy data

```{r, echo=TRUE, eval=TRUE}
Gestation_igwa <- select(Gestation, id, gestation, wt, age) 
Gestation_long <- pivot_longer(   data            = Gestation_igwa,   
                                  cols            = c(gestation, wt, age),    
                                  names_to        = "name",    
                                  values_to       = "value") 
head(Gestation_long, 6)
```

-   NB: we need to use "" quotes for `names_to` and `values_to` arguments because they are strings defining new columns

## Long and wide tidy data

```{r, echo=TRUE, eval=TRUE}
Gestation_long <- pivot_longer(Gestation_igwa, -id) 
head(Gestation_long, 6) 
```

Or specify which columns not to pivot, e.g. `-id` selects all variables except `id`

## Long and wide tidy data {.smaller}

-   To convert to a wider format, we use `pivot_wider`
-   For example, we specify:
    -   the data source
    -   where we get the new column names from
    -   where we get the new column values from

```{r, echo=TRUE, eval=TRUE}
Gestation_wide <- pivot_wider(data = Gestation_long,          
                              names_from  = name,
                              values_from = value) 

head(Gestation_wide, 3) 
```

## Pipe {.smaller}

-   dplyr imports the `%\>%` pipe from magrittr

-   `f(g(x))` is equivalent to `x %\>% g %\>% f`

-   Makes it easier to chain operations together without storing temporary objects

-   Output on left of `%\>%` becomes first argument of function on right

-   by convention, all tidyverse functions take a data frame as their first argument

```{r, echo=TRUE, eval = FALSE}
x %>% f_1 %>% f_2 %>% f_3
# rather than f_3(f_2(f_1(x)))
# or even worse... 
x_1 <- f_1(x) 
x_2 <- f_2(x_1) 
x_3 <- f_3(x_2)
```

## Pipe {.smaller}

-   An example

```{r, echo=TRUE, eval=TRUE}
Gestation %>% 
  group_by(race, smoke) %>% 
  summarise(wt = mean(wt)) %>% 
  pivot_wider(names_from  = race,
              values_from = wt)
```

-   NB the smoke variable is character and sorted alphabetically

# Summary

## Summary

-   Summarising data
    -   `group_by` to set a group structure
    -   `summarise` to calculate summary stats across group structure
    -   `count` to see how many rows in each group
    -   Reshaping data frames
        -   `pivot_longer` from variables side by side to key-value
        -   `pivot_wider` from key-value to named column variables

## Summary

-   Dealing with rows and columns
    -   `mutate` to create/modify/delete columns
    -   `select` to choose columns
    -   `filter` to choose rows based on logical condition
    -   `slice\*` to choose rows based on position or property
-   Pipe
    -   %\>% to chain operations
-   Wickham (2014) on what tidy data is
-   Wickham et al. (2019) for more explanation of tidyverse

# Visualisation with ggplot2

## Why do we visualise?

"Since the aim of exploratory data analysis is to learn what seems to be, it should be no surprise that pictures play a vital role in doing it well. There is nothing better than a picture for making you think of questions you had forgotten to ask (even mentally)." 

Tukey and Tukey (1985)

## Principles

Tufte (1983) and Pantoliano (2012) 

-   Show the data 
-   Provide clarity 
-   Allow comparison where appropriate 
-   use aesthetics to draw attention to important details 
-   make clear that data has multiple levels of structure

## Principles

-   Produce graphs with high data density
    -   make every drop of ink count
    -   careful use of whitespace
-   Avoid excessive and unnecessary use of graphical effects
-   Reader should be able to understand what the graph means and not be
    -   misled into thinking something that is untrue
    -   distracted from the main point

## Building plots

-   ggplot2 uses a grammar of graphics (Wickham 2010)
-   map variables in data frame to aesthetic options in the plot
-   choose a geometry for how to display these variables
-   adjustments to axis scales
-   adjustments to colors, themes, etc.
-   adding extra commands in a 'do this, then do this' manner
-   python users have plotnine (Kibirige 2020) which is based on the same ideas

## Building plots

How do we structure a call to ggplot to make a plot?

```{r, echo=TRUE, eval=FALSE}

# library(ggplot2) # already loaded with tidyverse 
ggplot(data = my.data.frame,
       aes(x = my.x.variable, 
           y = my.y.variable)) +   
  geom_point()
```

-   load ggplot2 package
-   Specify we want a ggplot object and which data frame we're going to use,
-   Set aesthetic options to map to the x and y axes of the plot
-   State geometry we're using to show variables

## Building plots {.smaller}

-   For example, consider daily maximum temperature varying with solar radiation in New York City 1973
-   Each row is a pair of values (x, y), shown as a point

```{r,eval=TRUE,echo=TRUE}

data(airquality) 
solar_temp_plot <- ggplot(data = airquality, aes(x = Solar.R, y = Temp)) + geom_point()  
solar_temp_plot
```

## Scatter plot {.smaller}

-   We can add some human-friendly labels and change the theme

```{r}

solar_temp_plot <- solar_temp_plot + theme_bw() + 
  labs(x = 'Solar radiation (Langleys)', y = 'Max. daily temp. (F)')  
solar_temp_plot
```

# Some more geometries

## Line plot {.smaller}

-   Similar to scatter plot, but joins pairs of values
-   Useful when showing how something changes over time
-   If (x, y) pairs ordered by
-   x, use geom_line() (e.g. x is time)
-   row order, use geom_path()
-   nothing, don't use a line

```{r,echo=T,eval=T}

airquality <- mutate(airquality, Date = as.Date(paste('1973', Month, Day, sep = '-'))) 
airquality_plot <- ggplot(data = airquality, aes(x=Date, y=Ozone)) +   
  geom_line() + 
  theme_bw() + 
  labs(y     = 'Ozone concentration (ppb)',                                   
       title = 'Daily mean Ozone in NYC (1973)')
```

## Line plot {.smaller}

```{r echo=TRUE, eval=TRUE}
airquality_plot
```

-   You may see this referred to as a time series plot

## Line plot {.smaller}

-   Observations whose neighbours are NA values can't be plotted with a line
-   Can layer multiple geometries for same aesthetic mapping

```{r, echo=TRUE, eval=TRUE}
airquality_plot + 
  geom_point()
```

## Scatterplot smoother {.smaller}

-   Often too much data in a scatter plot to see pattern - Maybe we want to highlight the trend in the data

```{r, echo=TRUE, eval=TRUE}
airquality_plot + 
  geom_point() + 
  geom_smooth() 
```

## Boxplot {.smaller}

-   continuous y, discrete x
-   outliers (\> 1.5 IQR from median) shown as points automatically


```{r, echo=TRUE, eval=TRUE}
ggplot(data = airquality, aes(x = factor(Month), y = Ozone)) + 
  geom_boxplot() + 
  theme_bw() + 
  labs(y = 'Ozone conc. (ppb)', x = 'Month')
```

## Histograms {.smaller}

-   univariate graphical summary needs only one aesthetic, x

```{r, echo=TRUE, eval=TRUE}
ozone_hist <- ggplot(data = airquality, aes(x = Ozone)) +     
  geom_histogram(binwidth = 10, boundary = 0) +     
  labs(x = 'Ozone concentration (ppb)') + theme_bw() 
ozone_hist
```

## Density plots {.smaller}

-   Kernel smoothing (continuous analogue of histogram)

```{r, echo=TRUE,eval=TRUE}
ozone_dens <- ggplot(data = airquality, aes(x = Ozone)) +     
  geom_density(fill = 'grey35') +     
  labs(x = 'Ozone concentration (ppb)') + 
  theme_bw() 
ozone_dens
```

# More on aesthetics

## More on aesthetics {.smaller}

::::{.columns}

:::{.column width="50%"}
| Aesthetic | What it affects             |
|-----------|-----------------------------|
| size      | points, lines               |
| shape     | points                      |
| linetype  | lines                       |
| colour    | points, lines, boundary     |
| alpha     | transparency                |
| fill      | interior                    |
| group     | recreates geometry by group |
:::

:::{.column width="50%"}
-   If these (except group) are outside aes() they fix the value for all parts of that geometry
-   Aesthetics specified inside ggplot() are inherited by all geometries for that plot
-   Not all geometries accept all aesthetics (e.g. geom_line() has no fill)
-   Some point shapes admit a colour and a fill
:::

::::

## More on aesthetics

```{r, echo=TRUE, eval=TRUE}
data(airquality) 
solar_temp_plot_colored <-    ggplot(data = airquality,        
                                     aes(x = Solar.R, y = Temp)) +   
  geom_point(aes(fill = factor(Month)),              
             shape = 21,              
             color = 'black')  +   
  labs(x = 'Solar radiation (Langleys)',        
       y = 'Max. daily temp. (F)') +   
  theme_bw() +    
  scale_fill_brewer(palette = "Purples",                       
                    name   = 'Month')
```

## More on aesthetics

```{r, echo=TRUE, eval=TRUE}
solar_temp_plot_colored 
```

# Small multiples

## Small multiples

-   Group a plot by some categorical variable
-   Repeat a basic graph for groups in the data
-   air quality data has information about, e.g. months
-   Can view 3-5 dimensions in the data on a 2D page
-   Often a better alternative to 3D, since it doesn't distort comparisons
-   Inner axes relate to the smallest X-Y plots
-   Outer axes relate to the grouping variables
-   Avoids writing loops

## Small multiples {.smaller}

-   Repeat histogram for each value of Month, one per facet

```{r, echo=TRUE, eval=TRUE}

ozone_hist + facet_wrap( ~ Month)
```

## Small multiples {.smaller}

We can also use facet_grid() to repeat the aesthetic and geometries for specified rows and cols variables

```{r, echo=TRUE, eval=TRUE}
library(gapminder) 
data(gapminder)

gapminder_plot <- ggplot(data = subset(gapminder, year >= 1992),        
                            aes(x = gdpPercap/1e3,
                                y = lifeExp)) +      
  geom_point(shape = 1, size = 0.5) +     
  facet_grid(rows = vars(year),                 
             cols = vars(continent)) +     
  scale_x_log10(labels = ~sprintf("%g", .)) +   
  xlab("GDP per capita ($k)") +   
  ylab("Life expectancy at birth (years)") +   
  theme_bw() +   
  theme(panel.grid.minor.x = element_blank())
```

## Small multiples {.smaller}

```{r, echo=TRUE, eval=TRUE}
gapminder_plot
```

# Summary

## Summary

-   We make graphs to tell a story with data
-   Should draw reader in and explain what they're seeing
-   Plots are built from
-   geometric objects
-   axis scales
-   coordinate systems (linear or logarithmic scale, 2D, 3D, etc.)
-   annotations (e.g. heading in small multiples) 

## Summary
-   Successively building a plot with a grammar of graphics allows development of complex plots from simple elements and small changes
-   Choose a plotting geometry that helps tell the story
-   Meaningful labels remove ambiguity and confusion
-   Be careful not to put too much in

## Further reading {.smaller}

The #r4ds community have TidyTuesday which makes use of the ideas in Wickham and Grolemund (2017)

-   History of visualisation
-   Friendly (2005)
-   Friendly (2006)
-   Visualisation to help decision making - Tufte (1997)
-   ggplot2 resources
-   RStudio (2021)
-   Chang (2017)

## References {.smaller}

Chang, Winston. 2017. R Graphics Cookbook: Practical Recipes for Visualizing Data. 2nd ed. O'Reilly Media. http://www.cookbook-r.com/Graphs/.

Friendly, M. 2005. "Milestones in the History of Data Visualization: A Case Study in Statistical Historiography." In Classification: The Ubiquitous Challenge, edited by C. Weihs and W. Gaul, 34--52. New York: Springer. http://www.math.yorku.ca/SCS/Papers/gfkl.pdf.

---------. 2006. "A Brief History of Data Visualization." In Handbook of Computational Statistics: Data Visualization, edited by C. Chen, W. Härdle, and A Unwin. Vol. III. Heidelberg: Springer-Verlag. http://www.datavis.ca/papers/hbook.pdf.

Kibirige, Hassan. 2020. A Grammar of Graphics for Python. https://plotnine.readthedocs.io/en/stable/index.html.

Nolan, Deborah, and Terry P Speed. 2001. Stat Labs: Mathematical Statistics Through Applications. Springer Science & Business Media. Pantoliano, Mike. 2012. "Data Visualization Principles: Lessons from Tufte." 2012. https://moz.com/blog/data-visualization-principles-lessons-fromtufte
