---
title: "Exploratory Data Analysis of Gestation data"
author: "Your Name"
date: "2024-02-23"
output: html_document
---

# Exploratory Data Analysis of Gestation data

```{r setup, include=FALSE}
library(tidyverse)
library(mosaicData)
# if you don't have mosaicData, install it with install.packages("mosaicData")
```

```{r read_data, echo=TRUE}
# read the data in
data(Gestation)
```

## Activity 1 - Quick look at the data
Calculate the number of observations in the data set, then calculate the number of observations per racial group and per racial group and level of mother's education.

```{r activity1, echo=TRUE}
# number of observations
count(Gestation)

count(Gestation, ...)

Gestation_n_race_ed <- count(Gestation, ...)
```

## Activity 2 - Further summary statistics
Calculate the mean age of mothers across all births. Ensure you use a human friendly name for the value you're creating.

Then calculate both mothers' mean age and babies' mean weight

```{r activity2, echo=TRUE}
mean_age <- summarise(Gestation, `Mean age` = ...)


summarise(Gestation, 
          `Mean age` = ...,
          `Mean wt`  = ...)
```

## Activity 3 - Grouped summaries
Make a new data frame containing only id, age and race variables, then calculate the mean age of mothers by race.
```{r activity3, echo=TRUE}
age_race <- select(Gestation, ...)

mean_age_by_race <- Gestation %>%
  group_by(...) %>%
  summarise(`Mean Age` = ...)
```

## Activity 4 - Extensions

### Activity 4a - Correlation
Calculate the correlation between age and weight across all births, then calculate the correlation between age and weight for each race.
```{r activity4a, echo=TRUE}
correlation_all <- cor(...)

correlation_by_race <- Gestation %>%
  group_by(...) %>%
  summarise(correlation = cor(...))
```

### Activity 4b - Multiple summary statistics
Calculate the sample mean of the ages and weights of the mothers in each race group
```{r activity4b, echo=TRUE}
mean_by_race <- Gestation %>%
  group_by(race) %>%
  summarise(mean_age = ..., mean_weight = ...)
```

### Activity 4c - Pivoting wider
Make a wide table from the summary data frame calculated in Activity 1 that has the number of observations for each combination of mother's education level and race.
Make each row is an education level and each column a race group.

```{r activity4c, echo=TRUE}
wide_summary <- Gestation_n_race_ed %>%
  pivot_wider(names_from = ..., values_from = ..., values_fill = 0)
```

### Activity 4d - Multiple summary statistics
Calculate the mean, standard deviation, minimum, maximum and proportion of values missing for the mothers' ages for each race group.
Hint: you *can* use `summarise_at()` for this but you could also just `summarise()`
```{r activity4d, echo=TRUE}

summary_stats_age <- Gestation %>%
  group_by(race) %>%
  summarise(mean_age = ...,
            sd_age = ...,
            min_age = ...,
            max_age = ...,
            prop_missing = ...)
```